---
title: "StatR 510 HW 1 Solutions"
output:
  pdf_document:
    fig_height: 4
    fig_width: 6
    toc: no
  html_document:
    toc: no
---

# 4.

## (a)

This is exactly what the `seq` function is meant to do.

```{r}
seq(from = 2, to = 100, by = 2)
```

## (b)

The `:` operator is convenient shorthand when your sequence increases by 1.
Multiplying by `2` gives us all even numbers.

```{r}
(1:50) * 2
```

## (c)

The `rep` function *repeats* the first argument the specified number of times.
The `cumsum` function takes the cumulative sum of the resulting vector.

```{r}
cumsum(rep(2, 50))
```

# 5.

Let's create a vector on which to operate. Yours probably looked different, but
the ideas are the same.

```{r}
x <- 1:50
```

## (a)

$\sum_{i=1}^n X_i$

```{r}
sum(x)
```

## (b)

$\mathrm{SS} = \sum_{i=1}^n X_i^2$

```{r}
sum(x^2)
```

## (c)

$\overline{X} = \frac{1}{n} \sum_{i=1}^n X_i$

```{r}
mean(x)
```

## (d)

$\sum_{i=1}^n (X_i - \overline{X})^2$

```{r}
sum((x - mean(x))^2)
```

## (e)

$\mathrm{SS} - n \overline{X}^2$

```{r}
sum(x^2) - length(x) * mean(x)^2
```

## Optional proof

\begin{align*}
\sum_{i=1}^n(X_i-\overline{X})^2 & =\sum_{i=1}^n\left(X_i-\overline{X}\right)\left(X_i-\overline{X}\right)\\
& = \sum_{i=1}^n\left(X_i^2-2X_i\overline{X}+\overline{X}^2\right)\\
& = \sum_{i=1}^nX_i^2 - 2\overline{X}\sum_{i=1}^nX_i + \sum_{i=1}^n\overline{X}^2\\
& = \sum_{i=1}^nX_i^2 - 2\overline{X}n\overline{X} + n\overline{X}^2 \qquad\qquad \left(\mbox{ since } \sum_{i=1}^n1 = n\right)\\
& = \sum_{i=1}^nX_i^2 - n\overline{X}^2\\
& = SS - n\overline{X}^2.
\end{align*}

# 6.

In setting this problem up, I use the `set.seed()` command below. This is a way
to make random number generation reproducible. This way, you can get the same
results I did below by using the same seed.

```{r,tidy=TRUE}
set.seed(1)

Names <- c("Alana", "Bettie", "Consuela", "Dona", "Elaine", "Frances", "Gerri",
           "Helene", "Ichabod", "Jin", "Kenyatta", "Larry", "Mikhailo", "Nick",
           "Odin")
Sex <- c(rep("F", 8), rep("M", 7))
Grades <- round(runif(15, 50, 100))
Grades
```

You can see the 15 different grades generated from a uniform distribution between 50 and 100.

## (a) Which grades were greater than or equal to 90?

We use the square brackes `[` and `]` to *subset* the `Grades` vector.

```{r}
gt90 <- Grades[Grades >= 90]
gt90
```

## (b) Who earned the highest grades?

To extract just the names:

```{r}
Names[gt90]
```

## (c) Who earned the lowest grades?

We can use the expression `Grades < 60` inside the square brackets *without* first assigning it to a variable name as we did with `gt90`.

```{r}
Names[Grades < 60]
```

## (d) Extracting the grades of the male and female students

```{r}
Grades.M <- Grades[Sex == "M"]
Grades.F <- Grades[Sex == "F"]
Grades.F
```

## (e) Average grades of the male and female students

```{r}
sum(Grades.F) / length(Grades.F)
sum(Grades.M) / length(Grades.M)
```

Clearly the guys need to step up their game!

## Problem 8.

## (a)

The `ls` function returns a character vector of the names of all objects in our
global environment. To remove them all, we use that as `list` argument to the
`rm` function:

```{r}
rm(list = ls())
```

## (b)

When adding two vectors of different length, the shorter of the two will be
repeated as many times as necessary to complete the addition. Adding a vector of
length 8 to a vector of length 2, for example, will repeat the vector of length
2 four times. In the example given, what R is really doing is `c(0,2,0,2) +
c(3,4,5,6)`. I R this is called *recycling*.

## (c)

A `vector` can only contain objects of the same class, e.g., `character` or
`numeric`. A `list`, by contrast, can be a collection of just about anything: you
can construct a list where its first object is a vector, the second is a data
frame, and the third is another list.

## (d)

The `length` and `mode` functions return information *about* an object. Other
useful descriptive functions are `names`, `str`, and `class`.

## (e)

In the help file for a function, the "Value" section describes the output of the
function.
